UCD\\MSc Thesis Proposal:\\\emph{An approach for Continuous Capacity Planning in Cloud Environments with Uptime-based Pricing Model}

### 2015-04-26 paper/paper.tex
final version

\maketitle
\tableofcontents\pdfbookmark[0]{Table of Contents}{toc}\newpage
\listoffigures\newpage
\par{\textbf{List of tables}}

@@

\textsl{New Infrastructure as a Service solutions are becoming available with a growing number of supported pricing models. A hosted Cloud environment can be a choice to design and build an infrastructure for a product. The recent availability of different pricing schemes based on resource utilization and uptime reveals new challenges in an already complex capacity planning process. There is a choice between ad-hoc provisioning and time commitment that comes with reduced hourly rates. When one decides for a time commitment, there are still further pricing options available. Few different prices for the same resource that depend on a daily uptime. Which one is the best one? When exactly does one pricing scheme becomes more cost effective that the other? We attempted to answer these critical questions by i) aggregating and transforming metrics data, ii) designing key optimality indicators, iii) using the indicators to build algorithms that suggest better configuration options, iv) using an existing large dataset to evaluate the algorithms. }

\end{abstract}
\newpage

Thanks to my wife Karen for her constant support and patience during the research and writing
of this Thesis.
Also, thanks to Prof. John Murphy and Viliam Holub for their support, help and assistance.

@@

IaaS consumer can face a situation when resource-based capacity is known and it is time to rent an optimal machine configuration from the provider. It raises few open ended questions, including i) how many VMs is needed? ii) what kind of VMs and which pricing scheme? iii) how long should each VM run? In this thesis, we attempt to answer these questions in a cost-aware fashion considering uptime-based pricing models. It is a new consideration that adds uptime parameter to the standard resource-aware provisioning techniques~\cite{Bartolini:2014:AFC:2658949.2637480}.

@@

Most of the related work is conducted from an IaaS provider perspective~\cite{6274129}~\cite{6253563} or aimed to find an optimal pricing state for providers and consumers to collaborate~\cite{6676685}~\cite{6963393}. This thesis is different in a way that it analyses a SaaS provider capacity planning in an environment when one resource have different prices tightly coupled with an uptime.

@@

Our proposed solution implements listed objectives using mainly map-reduce jobs executed on Apache Spark\footnote{\url{https://spark.apache.org}} platform. Aggregations, key indicators and algorithms were designed, implemented in pseudo-code and Python and evaluated on Google cluster-usage trace~\cite{clusterdata:Reiss2011}.

@@

Main challenges include dealing with dynamic changes of pricing schemes offered by IaaS providers. Even though this thesis uses Amazon's related taxonomy, it should be read as an attempt to solve a general case for uptime-based variable pricing capacity planning. Other challenging aspect is a cost of migration to suggested environment. It was not considered in the results, however, it created an opportunity for future work to investigate an impact of the delta difference between an old and new configuration cost. Lastly, we hit a technical issues related to implementation and execution of aggregations and algorithms on large data sets. Jobs were failing due to not enough RAM capacity on a single machine. The issues were resolved by a combination of code improvements, like changing the way aggregations build hash-maps and an increase of parallelism and chunking granularity on the platform.

@@

\myparagraph{Structure of the Thesis}\\
\textbf{Background}. This section outlines the background information related to IaaS consumer point of view and challenges related to finding the best configuration for the money. \par
\textbf{Design and Implementation}. This section explains design and environmental conditions for the aggregations, key indicators and algorithms. It also dives into implementation details on the Apache Spark platform. \par
\textbf{Evaluation}. The proposed map-reduce jobs and algorithms are evaluated on Google cluster-usage trace data. \par
\textbf{Conclusion and Future Work}. The suggestion results data and future work related to algorithm improvements are presented.

@@

We look from the perspective of Software as a Service (SaaS) provider that uses an Infrastructure as a Service (IaaS) provider's services to host a product. \\
SaaS provider seeks i) predictable cost estimates and ii) the best value for money. The latter for obvious reasons - every healthy business is interested in minimizing the costs. The former is needed to set a right product pricing. Not too expensive and not too cheap. Failure to do so may ruin a business's budget. \\
IaaS provider pricing can be flexible as described in below sections. One instance configuration can have many prices based on a daily uptime. SaaS provider faces the challenge to pick the best one.

@@

\paragraph{On-demand instances.} Allow to pay by the hour with no other commitments. The most expensive option.
\paragraph{Spot instances.} A real-time bidding platform. A customer can bid on price. Instance can run only when the bid exceeds current Spot Price.
\paragraph{Reserved instances.} Discounted price in comparison to On-demand in exchange of 1 or 3 years commitment. Guaranteed availability of reserved capacity.

This thesis focuses around the concept of Reserved Instances. The scheme is further split into Light, Medium and Heavy options. They differ in upfront and hourly costs. Based on pricing data, Light serves well for short periods of daily usage, Heavy is optimized for long workloads and Medium is in between. \\
This scheme got simplified on December 2014~\cite{AWS:pricing_change} and naming used in this thesis does not match the latest offering anymore. However, the idea of having the same instance configuration priced differently based on uptime is still valid and is used by other IaaS providers, e.g. Google Compute Engine.

@@

Google had greatly simplified the Compute Engine pricing. There is no concept of reservation nor bidding. All of the options are priced within pay-as-you-go model with 1 minute granularity. However, Google offers increasing discounts for sustained usage - 20\% for 25-50\% of month usage, 50-75\% of usage yields 40\% and everything above 75\% is priced with 60\% of discount rate. This is similar to the concept of Reserved Instances in a sense that uptime value influences a final price for the instance.

@@

In the work conducted in paper~\cite{6676685}, authors proposed a two-stage provisioning scheme. At the first stage, they proposed a solution that works with flat and on-demand instances. Secondly, they studied bidding for the spot instances as a Stackelberg game. Although not focused on reserved instances, authors developed the system model that might serve as an inspiration for future work to provide suggestions based on application specific QoS metrics.

@@

\section{Taxonomy}
This section explains a terminology being used for further design and evaluation stages. It is important to re-define some terms, like cost per hour, which is sometimes presented by IaaS providers in a non-complete way and may not include all costs. We also define new terminology that will be used throughout this thesis.

@@

The section contains design and implementation details for an end-to-end approach that analyzes machine configuration and proposes changes.
We start with metrics data prerequisites and guide through data transformations. We further define key indicators and formalize their implementations. The indicators can be used to filter non-optimal configurations. Algorithms that take the filter data as an input and generate configuration suggestions are presented later.

@@

The following list of metrics is required.
\myparagraph{Machine Id}
A unique machine ID. Used as a key to compute further analytics.

@@

Metrics should be aggregated in \textless K, List\textless V\textgreater\textgreater format where K represents Machine Id and List of V is a list of measurements in a given period.

@@

Metrics gathered in Prerequisites section have limited knowledge about the environment. Collection per job or per process produces one log line and has a scope of a measurement period. Such structured data would not help to answer machine-wide nor cluster-wide questions. The data has to be post-processed to allow calculation of below-listed aggregations and analytics.

@@

To comply with IaaS pricing strategies, the data is analyzed in terms of a daily usage. This requires splitting metrics into buckets of one day worth of data. 
It is achieved by applying a simple Apache Spark filter presented in Listing~\ref{daily_usage} that calculates an integer to indicate a day number based on a metric timestamp. The implementation can vary for various timestamp formats. Results can be saved in separated folders or tables and all other aggregations and transformations are executed in scope of 1 day worth of data.

@@

This metric indicates a daily number of hours of machine being up. It is a key metric that is be used to calculate optimality indicators discussed in the next three following paragraphs. Algorithm~\ref{alg:uptime_per_machine} represents pseudo-code implementation and Listing~\ref{uptime_per_machine_implementation} contains implementation in Python with Apache Spark. The data is extracted from timestamp values previously grouped into daily data buckets. By design, this approach extracts start and end time of jobs recorded in metrics using map function (lines 1-2). Then, two reduce jobs are executed that find the earliest start time (lines 3-4) and the latest end time (lines 5-6). As a result, we get two timestamps that indicate a range of a daily uptime. This design is flawed in a sense that if there are only two jobs logged - one in the morning and one in the evening - it will assume that machine was up no matter what happened in between.

@@

This is an implementation of \textit{cph} metric~\ref{eq:cph} defined in Taxonomy. Algorithm~\ref{alg:cost_per_hour} represents pseudo-code and Listing~\ref{cost_per_hour} shows implementation in Python. 
The algorithm takes upfront cost, charge per hour and daily uptime as input parameters. It aggregates both costs and calculates a real (total) cost per hour~\textit{if} machine were to run every day for a number of previously specified hours. We use this algorithm to build lookup tables for a 24 hour usage period. Essentially, we pass the same pricing scheme parameters and specify daily uptime from 1 to 24. This creates a table with hours per day to~\textit{cph} mapping.

@@

At 18 hours, the point of scheme cost intersection (highlighted as 1), it does not matter which scheme is chosen as cost is the same. This is point 0.
At each side of the intersection point either scheme1 or scheme 2 is more preferable.
For example, a machine running for 20 hours on scheme1 is running 2 hours outside our optimum 18 hour scheme cost intersection. This gives us a score of -2 for this machine compared to the same machine running on scheme 2 where the benefits continue to increase the longer the machine is running.
The same is true for a machine running for 10 hours on scheme 2. We are running at -8 below our optimum. In this case we would be more cost beneficial to run this machine on scheme1. Giving us a score of +8.

@@

A user can define which KDI value is non-optimal and run a report to find machines below this point. For example, every negative KDI value signalizes there is a better pricing scheme available. A user can catch this situation by having a threshold value set to 0. However, there might be a situation when machine is taken for maintenance and reports negative values for one or a couple of days. To avoid this kind of noise, a user can add an another criteria to specify a minimum number of days of reported values below the threshold.

@@

The proposed strategy calculates an amount of spare CPU and memory capacity based on cluster-wide averages. An aggregated value for CPU and memory represents an amount of capacity gathered from machines that are utilized below the average. Based on the amount of an aggregated capacity, the algorithm attempts to squeeze in a maximum amount of machines returned by a threshold-based report. As a result, it proposes to turn off some number of machines that can fit in the previously calculated below-average capacity. \\
This approach has a potential to perform well in under-utilized clusters. For well-balanced environments with similar usage patterns, it does not find a lot of spare capacity and would propose fairly limited improvements. \\
A pseudo-code implementation of this idea is presented in Algorithm~\ref{alg:squeeze-them-in}.

\begin{algorithm}[h]
\caption{Squeeze them in suggestion strategy}
\label{alg:squeeze-them-in}
 \KwData{Cluster averages for CPU and memory, List of machines from Threshold Based Report, List of other machines}
 \KwResult{List of machines that can be powered down}
 \algrenewcommand\algorithmicfunction{\textbf{class}}
 \algrenewcommand\algorithmicprocedure{\textbf{method}}
 \begin{algorithmic}[1]
  $float spareCpu = 0;$
  $float spareMemory = 0;$
  $ID[ ] machinesToExclude;$

\ForAll{machine in otherMachines}{
        \If{$machine.cpu < cluster.cpu$} {
                 $spareCpu += clusterCpu - machineCpu$
        }
        \If{$machine.memory < cluster.memory$} {
        	$spareMemory += clusterMemory - machineMemory$
        }
}

\ForAll{machine in machinesFromReport}{
        \If{$(spareCpu - machine.cpu > 0) \&\& (spareMemory - machine.memory > 0)$}{
                add machine to machinesToExclude;
        }
}

\Return machinesToExclude;
\end{algorithmic}
\end{algorithm}

@@

New configuration is calculated based on input from the report, list of available schemes and total uptime to handle. Pseudo-code implementation is presented in Algorithm~\ref{alg:new_configuration}. The algorithm iterates through all of available schemes, reads \textit{intersection points} and gets \textit{cph} for each of the \textit{ip}s. It returns a total cost and number of machines required for each proposed configuration. \\
Suggestion results are limited to one type of pricing scheme. In reality, there might be a more optimal solution yielded from mixing pricing schemes.

\begin{algorithm}[h]
\caption{New configuration suggestion}
\label{alg:new_configuration}
 \KwData{List of machines from Threshold Based Report, List of available schemes, Total uptime;}
 \KwResult{New configuration suggestion}
 \algrenewcommand\algorithmicfunction{\textbf{class}}
 \algrenewcommand\algorithmicprocedure{\textbf{method}}

  $float[ ] totalCost;$

\ForAll{scheme in availableSchemes}{
		\ForAll{ip in scheme.intersectionPoints}{
			vms = Uptime / ip.getHours() \\
			cost = ip.getCphForHours(ip.getHours()) \\
			totalCost[ip] = (vms, cost)
		}
}

\Return{totalCost}
\end{algorithm}

@@

Evaluation process can be split into four phases:
\begin{enumerate}
\item collection and transformation of pricing and usage data,
\item calculation of key indicators (\textit{cph, ip(scheme1, scheme2), KDI}) and cluster characteristics (\textit{total cost, performance averages}), 
\item suggestion of changes based on Algorithm~\ref{alg:squeeze-them-in} and~\ref{alg:new_configuration},
\item validation of a positive correlation between \textit{KDI} and \textit{total cost} changes.
\end{enumerate}

@@

Amazon Web Services\footnote{\url{http://aws.amazon.com/}} offered variable pricing schemes based on uptime. Customer could choose an instance configuration and total cost would be calculated based on a number of hours of uptime per day.
The offering was split into three options:
\begin{enumerate}
\item light - low upfront and high per hour cost,
\item medium - medium upfront and medium per hour cost,
\item heavy - high upfront and low per hour cost.
\end{enumerate}

@@

Table~\ref{tab:new_schema} is a new schema produced from join of capacity metrics from machine events and aggregations computed from tasks events data.

@@

\myparagraph{Scheme1}
\begin{table}[h]
\begin{center}
    \begin{tabular}{| l | l | l | l |}
    \hline
    \textbf{Hour} & \textbf{Light (price/hour)} & \textbf{Medium (price/hour)} & \textbf{Heavy (price/hour)} \\
    \hline
	1&5.7313972602739724&11.898904109589042&14.059630136986302 \\
    \hline
	2&3.317698630136986&6.2194520547945205&7.21031506849315 \\
    \hline
	3&2.513132420091324&4.326301369863014&4.9272100456621 \\
    \hline
	4&2.110849315068493&3.3797260273972602&3.785657534246575 \\    
	\hline
  5&1.8694794520547946&2.811780821917808&3.1007260273972603 \\
  	\hline
  6&1.7085662100456622&2.433150684931507&2.64410502283105 \\
  	\hline 
  7&1.593628180039139&2.162700587084149&2.317947162426614 \\
  \hline 
  8&1.5074246575342467&1.9598630136986301&2.0733287671232876 \\
  \hline
  9&1.4403774733637749&1.8021004566210044&1.8830700152207003 \\
  \hline
  10&1.3867397260273973&1.6758904109589041&1.73086301369863 \\ 
  \hline
  11&1.342854296388543&1.5726276463262765&1.6063300124533002 \\ 
  \hline
  12&1.3062831050228312&1.4865753424657535&1.5025525114155251 \\
  \hline
  13&1.2753382507903057&1.4137618545837725&1.414740779768177 \\
  \hline
  14&1.2488140900195697&1.3513502935420745&1.3394735812133072 \\
  \hline
  15&1.2258264840182649&1.2972602739726027&1.2742420091324203 \\
  \hline
  16&1.2057123287671234&1.249931506849315&1.2171643835616437 \\
  \hline
  17&1.1879645447219984&1.208170829975826&1.1668017727639 \\
  \hline
  18&1.1721887366818875&1.1710502283105022&1.12203500761035 \\ 
  \hline
  19&1.1580735400144198&1.1378370583994233&1.0819805335255948 \\
  \hline
  20&1.1453698630136988&1.107945205479452&1.045931506849315 \\
  \hline
  21&1.1338760600130462&1.0809001956947162&1.0133157208088714 \\
  \hline
  22&1.1234271481942715&1.0563138231631384&0.9836650062266501 \\
  \hline
  23&1.1138868374032165&1.0338653960690887&0.9565926146515782 \\ 
  \hline
  24&1.1051415525114157&1.0132876712328769&0.9317762557077626 \\ 
    \hline
    \end{tabular}
\end{center}
\caption{cph results for cc2.8xlarge reserved instance}
\label{tab:cph:scheme1}
\end{table}

\myparagraph{Scheme2}

>>>>

\maketitle
\tableofcontents\pdfbookmark[0]{Table of Contents}{toc}\newpage
\listoffigures\newpage
\listoftables

@@

\textsl{New Infrastructure as a Service solutions are becoming available with a growing number of supported pricing models. A hosted Cloud environment is a strong choice to design and build an infrastructure for a product. The recent availability of different pricing schemes based on resource utilization and uptime reveals new challenges in an already complex capacity planning process. There is a choice between ad-hoc provisioning and time commitment that comes with reduced hourly rates. When one decides for a time commitment, there are still more pricing options available. Few different prices for the same resource that depend on a daily uptime. Which one is the best one? When exactly does one pricing scheme becomes more cost effective that the other? We attempted to answer these critical questions by i) aggregating and transforming metrics data, ii) designing key optimality indicators, iii) using the indicators to build algorithms that suggest better configuration options, iv) using an existing large dataset to evaluate the algorithms. }

\end{abstract}
\newpage
\chapter*{Acknowledgements}

I would like to thank Prof. Liam Murphy for help and guidance.

I thank Christina Thorpe for many insightful conversations during the development of the ideas in this thesis. And for her time that I should use more.

I also wish to express my gratitude to Alan for inspiration to do this research.

I owe more than thanks to my lovely Klaudia for her support and constant encouragement during the entire process.

And finally, but not least, thanks goes to my parents who have always motivated me to achieve more.

@@

IaaS consumer can face a situation when resource-based capacity is known and it is time to rent an optimal machine configuration from the provider. It raises few open-ended questions, including i) how many VMs is needed? ii) what kind of VMs and which pricing scheme? iii) how long should each VM run? In this thesis, we attempt to answer these questions in a cost-aware fashion considering uptime-based pricing models. It is a new consideration that adds uptime parameter to the standard resource-aware provisioning techniques~\cite{Bartolini:2014:AFC:2658949.2637480}.

@@

Most of the related work is conducted from an IaaS provider perspective~\cite{6274129}~\cite{6253563} or aimed to find an optimal pricing state for providers and consumers to collaborate~\cite{6676685}~\cite{6963393}. This thesis is different in a way that it analyses a SaaS provider capacity planning in an environment when one resource has different prices tightly coupled with an uptime.

@@

Our proposed solution implements the objectives using mainly map-reduce jobs executed on Apache Spark\footnote{\url{https://spark.apache.org}} platform. Aggregations, key indicators and algorithms were designed, implemented in pseudo-code and Python and evaluated on Google cluster-usage trace~\cite{clusterdata:Reiss2011}.

@@

Main challenges include dealing with dynamic changes of pricing schemes offered by IaaS providers. Even though this thesis uses Amazon's related taxonomy, it should be read as an attempt to solve a general case for uptime-based variable pricing capacity planning. Another challenging aspect is a cost of migration to suggested environment. It was not considered in the results, however, it created an opportunity for future work to investigate an impact of the delta difference between an old and new configuration cost. Lastly, we encountered technical issues related to implementation and execution of aggregations and algorithms on large data sets. Jobs were failing due to not enough of RAM capacity on a single machine. The issues were resolved by a combination of code improvements, like changing the way aggregations build hashmaps and an increase of parallelism and chunking granularity on the platform.

@@

\myparagraph{Structure of the Thesis}\\
\textbf{Background}. This section outlines the background information related to IaaS consumer's point of view and challenges related to finding the best configuration for the money. \par
\textbf{Design and Implementation}. This section explains the design and environmental conditions for the aggregations, key indicators and algorithms. It also dives into implementation details on the Apache Spark platform. \par
\textbf{Evaluation}. The proposed map-reduce jobs and algorithms are evaluated on Google cluster-usage trace data. \par
\textbf{Conclusion and Future Work}. The suggestion results data and future work related to algorithm improvements are presented.

@@

We look from the perspective of Software as a Service (SaaS) provider that uses an Infrastructure as a Service (IaaS) provider's services to host a product. \\
SaaS provider seeks i) predictable cost estimates and ii) the best value for money. The latter for obvious reasons - every healthy business is interested in minimizing the costs. The former is needed to set a right product pricing. Not too expensive and not too cheap. Failure to do so may ruin a business's budget. \\
IaaS provider's pricing can be flexible as described in below sections. One instance configuration can have many prices based on a daily uptime. SaaS provider faces the challenge to pick the best one.

@@

\paragraph{On-demand instances.} Allow to pay by the hour with no other commitments. The most expensive option.
\paragraph{Spot instances.} A real-time bidding platform. A customer can bid on the price. An instance can run only when the bid exceeds current Spot Price.
\paragraph{Reserved instances.} Discounted price in comparison to On-demand in exchange for 1 or 3 years of commitment. Guaranteed availability of reserved capacity.

This thesis focuses around the concept of Reserved Instances. The scheme is split further into Light, Medium and Heavy options. They differ in upfront and hourly costs. Based on pricing data, Light serves well for short periods of daily usage, Heavy is optimized for long workloads and Medium is in between. \\
This scheme got simplified on December 2014~\cite{AWS:pricing_change} and naming used in this thesis does not match the latest offering anymore. However, the idea of having the same instance configuration priced differently based on uptime is still valid and is used by other IaaS providers, e.g. Google Compute Engine.

@@

Google had greatly simplified the Compute Engine pricing. There is no concept of reservation nor bidding. All of the options are priced within pay-as-you-go model with 1-minute granularity. However, Google offers increasing discounts for sustained usage - 20\% for 25-50\% of month usage, 50-75\% of usage yields 40\% and everything above 75\% is priced with 60\% of discount rate. This is similar to the concept of Reserved Instances in a sense that uptime value influences a final price for the instance.

@@

In the work conducted in paper~\cite{6676685}, authors proposed a two-stage provisioning scheme. At the first stage, they proposed a solution that works with flat and on-demand instances. Secondly, they studied bidding for the spot instances as a Stackelberg game. Although not focused on reserved instances, authors developed the system model that might serve as an inspiration for future work to provide suggestions based on application-specific QoS metrics.

@@

\section{Taxonomy}
This section explains a terminology being used for further design and evaluation stages. It is important to redefine some terms, like cost per hour, which is sometimes presented by IaaS providers in a non-complete way and may not include all costs. We also define new terminology that will be used throughout this thesis.

@@

This section contains design and implementation details for an end-to-end approach that analyzes machine configuration and proposes changes.
We start with metrics data prerequisites and guide through data transformations. We further define key indicators and formalize their implementations. The indicators can be used to filter non-optimal configurations. Finally, we present algorithms that take the filter data as an input and generate configuration suggestions.

@@

The following list presents required metrics.
\myparagraph{Machine Id}
A unique machine ID. Used as a key to compute further analytics.

@@

Metrics should be aggregated in \textless K, List\textless V\textgreater\textgreater format where K represents Machine Id and List of V is a list of measurements for a given period.

@@

Metrics gathered in Prerequisites Section have limited knowledge about the environment. Collection per job or per process produces one log line and has a scope of a measurement period. Such structured data would not help to answer machine-wide nor cluster-wide questions. The data has to be post-processed to allow calculation of below-listed aggregations and analytics.

@@

To comply with IaaS pricing strategies, the data is analyzed in terms of a daily usage. This requires splitting metrics into buckets of one day worth of data. 
It is achieved by applying a simple Apache Spark filter presented in Listing~\ref{daily_usage} that calculates an integer to indicate a day number based on a metric timestamp. The implementation can vary for various timestamp formats. Results can be saved in separated folders or tables and all other aggregations and transformations are executed in the scope of 1 day worth of data.

@@

This metric indicates a daily number of hours of a machine being up. It is a key metric that is be used to calculate optimality indicators discussed in the next three following paragraphs. Algorithm~\ref{alg:uptime_per_machine} represents pseudo-code implementation and Listing~\ref{uptime_per_machine_implementation} contains implementation in Python with Apache Spark. The data is extracted from timestamp values previously grouped into daily data buckets. By design, this approach extracts start and end time of jobs recorded in metrics using map function (lines 1-2). Then, two reduce jobs are executed that find the earliest start time (lines 3-4) and the latest end time (lines 5-6). As a result, we get two timestamps that indicate a range of a daily uptime. This design is flawed in a sense that if there are only two jobs logged - one in the morning and one in the evening - it will assume that a machine was up no matter what happened in between.

@@

This is an implementation of \textit{cph} metric~\ref{eq:cph} defined in Taxonomy. Algorithm~\ref{alg:cost_per_hour} represents pseudo-code and Listing~\ref{cost_per_hour} shows implementation in Python. 
The algorithm takes upfront cost, charge per hour and daily uptime as input parameters. It aggregates both costs and calculates a real (total) cost per hour~\textit{if} a machine were to run every day for a number of previously specified hours. We use this algorithm to build lookup tables for a 24-hour usage period. Essentially, we pass the same pricing scheme parameters and specify daily uptime from 1 to 24. This creates a set with hours per day to~\textit{cph} mapping.

@@

At 18 hours, the point of scheme cost intersection (highlighted as 1), it does not matter which scheme is chosen as the cost is the same. This is point 0.
At each side of the intersection point either scheme1 or scheme 2 is more preferable.
For example, a machine running for 20 hours on scheme1 is running 2 hours outside our optimum 18-hour scheme cost intersection. This gives us a score of -2 for this machine compared to the same machine running on scheme 2 where the benefits continue to increase the longer the machine is running.
The same is true for a machine running for 10 hours on scheme 2. We are running at -8 below our optimum. In this case, we would be more cost beneficial to run this machine on scheme1. Giving us a score of +8.

@@

A user can define which KDI value is non-optimal and run a report to find machines below this point. For example, every negative KDI value signalizes there is a better pricing scheme available. A user can catch this situation by having a threshold value set to 0. However, there might be a situation when a machine is taken for maintenance and reports negative values for one or a couple of days. To avoid this kind of background noise, a user can add an another criterion to specify a minimum number of days of reported values below the threshold.

@@

The proposed strategy calculates an amount of spare CPU and memory capacity based on cluster-wide averages. An aggregated value for CPU and memory represents an amount of capacity gathered from machines that are utilized below the average. Based on the amount of an aggregated capacity, the algorithm attempts to squeeze in a maximum amount of machines returned by a threshold-based report. As a result, it proposes to turn off some number of machines that can fit in the previously calculated below-average capacity. \\
This approach has a potential to perform well in under-utilized clusters. For well-balanced environments with similar usage patterns, it does not find a lot of spare capacities and would propose fairly limited improvements. \\
A pseudo-code implementation of this idea is presented in Algorithm~\ref{alg:squeeze-them-in}.

\begin{algorithm}[h]
\caption{Squeeze them in suggestion strategy}
\label{alg:squeeze-them-in}
 \KwData{Cluster averages for CPU and memory, List of machines from Threshold Based Report, List of other machines}
 \KwResult{List of machines that can be powered down}
 \begin{algorithmic}[1]
  \State $float spareCpu = 0;$
  \State $float spareMemory = 0;$
  \State $ID[ ] machinesToExclude;$
  \State \ForAll{machine in otherMachines}{
   \If {$machine.cpu < cluster.cpu$}{
  	 $spareCpu += clusterCpu - machineCpu$
	}
	\If{$machine.memory < cluster.memory$} {
	 $spareMemory += clusterMemory - machineMemory$
		}
	}
  \State \ForAll{machine in machinesFromReport}{
	  \If{$(spareCpu - machine.cpu > 0) \&\& (spareMemory - machine.memory > 0)$}{
	                $add machine to machinesToExclude;$
	        }
	}
\State\Return machinesToExclude;
\end{algorithmic}
\end{algorithm}

The algorithm takes CPU and memory cluster averages, a list of machines to optimize and list of the other machines as an input. First, it calculates the amount of spare capacity based on other machines data (code block 4). The results are saved in memory using variables listed in line 1 and 2. Then, it iterates through all of the machines that were identified to be non-optimal (code block 5) and if there is enough spare capacity it adds them to the exclusion list (line 3). Finally, it returns a list of machines that can be powered off.

@@

The new configuration is calculated based on input from the report, list of available schemes and total uptime to handle. Pseudo-code implementation is presented in Algorithm~\ref{alg:new_configuration}. The algorithm iterates through all of the available schemes, reads \textit{intersection points} and gets \textit{cph} for each of the \textit{ip}s. It returns a total cost and number of machines required for each proposed configuration. \\
Suggestion results are limited to one type of pricing scheme. In reality, there might be a more optimal solution yielded from mixing pricing schemes.

\begin{algorithm}[h]
\caption{New configuration suggestion}
\label{alg:new_configuration}
 \KwData{List of machines from Threshold Based Report, List of available schemes, Total uptime;}
 \KwResult{New configuration suggestion}
 \begin{algorithmic}[1]
  \State $float\{\} totalCost;$
  \State\ForAll{scheme in availableSchemes}{
		\ForAll{ip in scheme.intersectionPoints}{
			$vms = Uptime / ip.getHours()$ \par
			$cost = ip.getCphForHours(ip.getHours())$ \par
			$totalCost[ip] = (vms, cost)$\par
		}
}
\State \Return{totalCost}
\end{algorithmic}
\end{algorithm}

The algorithm takes a list of machines from the report, list of available schemes and total uptime to accommodate. It iterates through available schemes (code block 2) and for each of them attempts to suggest a configuration for a given uptime. The suggestion process is split into phases 1) calculate a number of VMs required to handle requested uptime -  VM's uptime time depends on \textit{intersection point} value, 2) calculate cost of running the VM based on a corresponding \textit{cph} value, 3) add results to totalCost map (line 1). Finally, the algorithm returns a map of total cost and VMs required for each analyzed scheme.

@@

Evaluation process can be split into four phases:
\begin{enumerate}
\item collection and transformation of pricing and usage data,
\item calculation of key indicators (\textit{cph, ip(scheme1, scheme2), KDI}) and cluster characteristics (\textit{total cost, performance averages}), 
\item suggestion of changes based on Algorithm~\ref{alg:squeeze-them-in} and~\ref{alg:new_configuration},
\item validation of a correlation between \textit{KDI} and \textit{total cost} changes.
\end{enumerate}

@@

Amazon Web Services\footnote{\url{http://aws.amazon.com/}} offered variable pricing schemes based on uptime. Customers could choose an instance configuration and the total cost would be calculated based on a number of hours of uptime per day.
The offering was split into three options:
\begin{enumerate}
\item light - low upfront and high per hour cost,
\item medium - medium upfront and medium per hour cost,
\item heavy - high upfront and low per hour cost.
\end{enumerate}

@@

Table~\ref{tab:new_schema} is a new schema produced from a join of capacity metrics from machine events and aggregations computed from tasks events data.

@@

\begin{table}[H]
\begin{center}
    \begin{tabular}{| l | l | l | l |}
    \hline
    \textbf{Hour} & \textbf{Light (price/hour)} & \textbf{Medium (price/hour)} & \textbf{Heavy (price/hour)} \\
    \hline
	1&5.7313972602739724&11.898904109589042&14.059630136986302 \\
    \hline
	2&3.317698630136986&6.2194520547945205&7.21031506849315 \\
    \hline
	3&2.513132420091324&4.326301369863014&4.9272100456621 \\
    \hline
	4&2.110849315068493&3.3797260273972602&3.785657534246575 \\    
	\hline
  5&1.8694794520547946&2.811780821917808&3.1007260273972603 \\
  	\hline
  6&1.7085662100456622&2.433150684931507&2.64410502283105 \\
  	\hline 
  7&1.593628180039139&2.162700587084149&2.317947162426614 \\
  \hline 
  8&1.5074246575342467&1.9598630136986301&2.0733287671232876 \\
  \hline
  9&1.4403774733637749&1.8021004566210044&1.8830700152207003 \\
  \hline
  10&1.3867397260273973&1.6758904109589041&1.73086301369863 \\ 
  \hline
  11&1.342854296388543&1.5726276463262765&1.6063300124533002 \\ 
  \hline
  12&1.3062831050228312&1.4865753424657535&1.5025525114155251 \\
  \hline
  13&1.2753382507903057&1.4137618545837725&1.414740779768177 \\
  \hline
  14&1.2488140900195697&1.3513502935420745&1.3394735812133072 \\
  \hline
  15&1.2258264840182649&1.2972602739726027&1.2742420091324203 \\
  \hline
  16&1.2057123287671234&1.249931506849315&1.2171643835616437 \\
  \hline
  17&1.1879645447219984&1.208170829975826&1.1668017727639 \\
  \hline
  18&1.1721887366818875&1.1710502283105022&1.12203500761035 \\ 
  \hline
  19&1.1580735400144198&1.1378370583994233&1.0819805335255948 \\
  \hline
  20&1.1453698630136988&1.107945205479452&1.045931506849315 \\
  \hline
  21&1.1338760600130462&1.0809001956947162&1.0133157208088714 \\
  \hline
  22&1.1234271481942715&1.0563138231631384&0.9836650062266501 \\
  \hline
  23&1.1138868374032165&1.0338653960690887&0.9565926146515782 \\ 
  \hline
  24&1.1051415525114157&1.0132876712328769&0.9317762557077626 \\ 
    \hline
    \end{tabular}
\end{center}
\caption{cph results for cc2.8xlarge reserved instance}
\label{tab:cph:scheme1}
\end{table}

\begin{table}[H]
\begin{center}
    \begin{tabular}{| l | l | l | l |}
    \hline
    \textbf{Hour} & \textbf{Light (price/hour)} & \textbf{Medium (price/hour)} & \textbf{Heavy (price/hour)} \\
    \hline
1&3.656821917808219&7.654246575342467&8.918575342465754\\ 
 \hline 
2&2.1664109589041094&4.087123287671233&4.595287671232876\\ 
 \hline 
3&1.669607305936073&2.898082191780822&3.1541917808219178\\ 
 \hline 
4&1.4212054794520548&2.303561643835616&2.4336438356164383\\ 
 \hline 
5&1.2721643835616436&1.9468493150684931&2.0013150684931507\\ 
 \hline 
6&1.1728036529680366&1.709041095890411&1.713095890410959\\ 
 \hline 
7&1.1018317025440314&1.5391780821917809&1.5072250489236791\\ 
 \hline 
8&1.0486027397260274&1.411780821917808&1.352821917808219\\ 
 \hline 
9&1.0072024353120244&1.3126940639269407&1.2327305936073059\\ 
 \hline 
10&0.9740821917808219&1.2334246575342467&1.1366575342465755\\ 
 \hline 
11&0.9469838107098382&1.1685678704856788&1.058052303860523\\ 
 \hline 
12&0.9244018264840183&1.1145205479452056&0.9925479452054796\\ 
 \hline 
13&0.9052939936775553&1.0687881981032665&0.9371211801896734\\ 
 \hline 
14&0.8889158512720158&1.0295890410958906&0.8896125244618396\\ 
 \hline 
15&0.8747214611872147&0.9956164383561644&0.8484383561643836\\ 
 \hline 
16&0.8623013698630138&0.9658904109589042&0.8124109589041095\\ 
 \hline 
17&0.8513424657534246&0.9396615632554393&0.7806220789685737\\ 
 \hline 
18&0.8416012176560121&0.9163470319634702&0.7523652968036529\\ 
 \hline 
19&0.8328853640951694&0.8954866618601299&0.7270829127613554\\ 
 \hline 
20&0.825041095890411&0.8767123287671232&0.7043287671232877\\ 
 \hline 
21&0.8179439008480104&0.8597260273972603&0.6837416829745597\\ 
 \hline 
22&0.8114919053549191&0.8442839352428394&0.6650261519302615\\ 
 \hline 
23&0.8056009529481835&0.8301846337105421&0.6479380583680763\\ 
 \hline 
24&0.8002009132420091&0.8172602739726028&0.6322739726027398 \\
 \hline 
    \end{tabular}
\end{center}
\caption{cph results for m2.4xlarge reserved instance}
\label{tab:cph:scheme2}
\end{table}
